Predicting Conflict Escalation

### 2023-05-02 docs/hes_final/article.qmd
Rearanged and renamed model sections

While many of GDELT’s databases reflect news articles from 1 January 1979 onward, GDELT’s news article embeddings exist only from 1 January 2020, limiting the time period for which training data can be drawn. Availability of ACLED’s monthly event and fatality totals varies by country, but event and fatality totals are available for all countries since at least 1 January 2020.
Our training data was derived from these GDELT and ACLED data. GDELT news article embeddings served as our independent variable, while the “spike” metric derived from ACLED’s conflict data served as our target variable. Hard dates were set for train, test, and validation splits, and none of our models saw data published after October 2022 during training, with the remaining months used for testing and validation.

@@

    GDELT --> Data(Time-partitioned train, val, & test datasets)
    Spike --> Data
```

### Statistical model using monthly embedding averages

## Feed Forward Neural Network

Our second approach employed a feed forward neural network model. Uniform manifold approximation and projection (UMAP) was used to reduce dimensionality of GDELT article embeddings in a supervised manner [@sainburg2021parametric]. Our spike variable was used to condition the dimension reduction process of the training data to encourage separation of the variable classes in the reduced embedding space. The fit UMAP model, was applied to the validation and test data without supervision. These reduced, 128 dimension embeddings, along with country context indicators, were used as inputs to a feed forward neural network with a single hidden layer of 32 neurons. The model's 4577 parameters were trained for 3 epochs.

@@

### Transformer Model using Resampled Embeddings

>>>>

### Training Dataset

While many of GDELT’s databases reflect news articles from 1 January 1979 onward, GDELT’s news article embeddings exist only from 1 January 2020, limiting the time period for which training data can be drawn. Availability of ACLED’s monthly event and fatality totals varies by country, but event and fatality totals are available for all countries since at least 1 January 2020.

Our training data was derived from these GDELT and ACLED data. GDELT news article embeddings served as our independent variable, while the “spike” metric derived from ACLED’s conflict data served as our target variable. The spike metric was used to label article data on a three-month time lag, allowing us to make predictions of future events. Hard dates were set for train, test, and validation splits, and none of our models saw data published after October 2022 during training, with the remaining months used for testing and validation.

@@

    GDELT --> Data(Time-partitioned train, val, & test datasets)
    Spike --> Data
```
## Conflict Prediction Using Text and Country Features

Our first approach employed a feed forward neural network model. Uniform manifold approximation and projection (UMAP) was used to reduce dimensionality of GDELT article embeddings in a supervised manner [@sainburg2021parametric]. Our spike variable was used to condition the dimension reduction process of the training data to encourage separation of the variable classes in the reduced embedding space. The fit UMAP model, was applied to the validation and test data without supervision. These reduced, 128 dimension embeddings, along with country context indicators, were used as inputs to a feed forward neural network with a single hidden layer of 32 neurons. The model's 4577 parameters were trained for 3 epochs.

@@

### Conflict Prediction Using Text and Country Labels

### Conflict Prediction Using Text Alone
